Theoretical aspects in gravity modelling

### 2018-04-22 manuscript/letter.tex

\title[Theoretical aspects in gravity modeling]
      {Theoretical aspects in gravity modeling}

\author[Oliveira Jr, Barbosa and Uieda]
{Vanderlei C. Oliveira Jr$^1$\thanks{Sei la, Nacional, Brazil}, Val\'{e}ria C. F. Barbosa$^1$ and Leonardo Uieda$^2$ \\
	$^1$ Department of Geophysics, Observat\'{o}rio Nacional, Rio de Janeiro, Brazil \\
	$^2$ Department of Geology and Geophysics, University of Hawaii, Manoa, USA
}

@@

\begin{summary}
 Gravity anomalies have long been used by geophysicists for the 
 purpose of determining density distributions in subsurface.

 However, gravity anomalies

 In this paper, we discuss the fundamental concepts

 URGENTE: \citep{marussi1974}, \citep{torge2012}, section 4.2.1
\end{summary}

\begin{keywords}
 potential fields -- gravity disturbance -- gravity anomaly -- gravity modeling.
\end{keywords}

@@

The resultant of gravitational force and centrifugal force acting 
on a body at rest on the Earth's surface is called 
\textit{gravity vector} and its intensity is called 
\textit{gravity} \citep{heiskanen-moritz1967,
hofmann-wellenhof-moritz2005}.
In the case of gravimetry on moving platforms (e.g., airplanes,
helicopters, marine vessels), there are additional
non-gravitational accelerations due to the vehicle motion, 
such as Coriolis acceleration and high-frequency vibrations 
\citep{glennie-etal2000,nabighian-etal2005-grav,baumann-etal2012}.
Geophysicists use gravity for estimating the Earth's 
internal density distribution whereas geodesists use
gravity to estimate the geoid \citep{li2001}.
Hence, geophysicists are usually interested 
in the gravitational component of the observed gravity, 
which is produced by the Earth's internal 
density distribution.

The first step of the procedure for isolating this 
gravitational component consists in removing the non-gravitational 
effects due to the vehicle motion and also the time variations 
such as Earth tides, instrumental drift and barometric 
pressure changes, for example.
If these effects are properly removed, the resultant 
gravity data can be considered as the sum of a 
centrifugal component due to the Earth's rotation and
a gravitational component produced by the whole Earth's
internal density distribution.
The isolation of this particular gravitational component 
and its subsequent use for estimating density 
distributions related to geological structures in subsurface 
are the main goals in applied geophysics \citep{blakely1996}.
The computation of the gravitational effect produced by
the geological structures is commonly known in geophysics as 
\textit{gravity modeling}.
This term is different from \textit{gravity field modeling}, which
is commonly used in geodesy to denote the characterization of the 
gravity field in a local, regional or global scale.

@@

\section{Terrestrial Reference Systems used in gravity modeling}

For geophysical purposes, there are three important Terrestrial Reference Systems used in gravity modeling. 
They rotate with the Earth and are used for describing
positions and movements of objects on and close to the Earth’s surface
\citep{torge2012}.

The first one is a geocentric system of Cartesian coordinates having 
the $Z$-axis coincident with the mean Earth's rotational axis,
the $X$-axis pointing to the Greenwich meridian and the $Y$-axis
directed so as to obtain a right-handed system (Figure \ref{fig:GCS-GGS}).
This reference system can be found in the literature with different names: Mean Terrestrial System \citep[e.g.,][]{soler1976}, Earth-fixed geocentric Cartesian system \citep[e.g.,][]{torge2012} 
or Earth-centered Earth-fixed system \citep[e.g.,][]{bouman_etal2013}, for example. Here, we opted for simply using the term Geocentric Cartesian System (GCS).

Other important reference system is a geocentric system of 
geodetic coordinates, which is defined by the reference ellipsoid 
used in the Normal Earth model \citep{heiskanen-moritz1967, soler1976, 
torge2012, bouman_etal2013}. 
In this coordinate system, the position of a point 
is defined by the \textit{geometric height} $h$, 
\textit{geodetic latitude} $\varphi$ and \textit{longitude} $\lambda$ (Figure \ref{fig:GCS-GGS}).
For convenience, we call this system Geocentric Geodetic System
(GGS). 
At a given point $(h, \varphi, \lambda)$, there are three 
mutually-orthogonal unit vectors (Figure \ref{fig:GCS-GGS}) given by \citep{soler1976}:
\begin{equation}
\begin{split}
\hat{\mitbf{u}} &= 
\begin{bmatrix}
\cos\varphi \, \cos\lambda \\
\cos\varphi \, \sin\lambda \\
\sin\varphi
\end{bmatrix} \\
\hat{\mitbf{v}} &= 
\begin{bmatrix}
-\sin\varphi \, \cos\lambda \\
-\sin\varphi \, \sin\lambda \\
\cos\varphi
\end{bmatrix} \\
\hat{\mitbf{w}} &= 
\begin{bmatrix}
-\sin\lambda \\
\cos\lambda \\
0
\end{bmatrix}
\end{split} \: .
\label{eq:unit-vectors}
\end{equation}
The required equations to convert coordinates $(h, \varphi, 
\lambda)$ referred to the GGS into coordinates $(X, Y, Z)$ referred to 
the GCS (Figure \ref{fig:GCS-GGS}) and vice versa can be easily found in 
the literature \citep[e.g.,][]{heiskanen-moritz1967, torge2012, 
bouman_etal2013}.

In a local or regional study, geophysicists commonly use a topocentric 
Cartesian coordinate system (TCS) with origin at a point $P$ on or 
close to the Earth's surface and axes $x$, $y$ and $z$ (Figure 
\ref{fig:TCS}). In the TCS, the axes $x$ and $y$ are parallel to 
the unit vectors $\hat{\mitbf{v}}_{P}$ $\hat{\mitbf{w}}_{P}$, respectively,
whereas the $z$-axis is opposite to the unit vector 
$\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) and points downward.
Consider, for example, a TCS (Figure \ref{fig:TCS}) with origin at a point $P$ with
coordinates $(X_{P}, Y_{P}, Z_{P})$ referred to the GCS (Figure \ref{fig:GCS-GGS}).
In this case, the relationship between the coordinates $(X, Y, Z)$ of a 
point in the GCS and the coordinates $(x, y, z)$ of the same point in the 
TCS is given by:
\begin{equation}
\begin{bmatrix}
x \\
y \\
z 
\end{bmatrix} =
\mitbf{R}^{\top} \begin{bmatrix}
X - X_{P} \\
Y - Y_{P} \\
Z - Z_{P}
\end{bmatrix} \: ,
\label{eq:GCS2TCS}
\end{equation}
where $\mitbf{R}$ is an orthogonal matrix whose columns
are defined by the unit vectors $\hat{\mitbf{u}}$, $\hat{\mitbf{v}}$ and
$\hat{\mitbf{w}}$ (equation \ref{eq:unit-vectors}) as follows:
\begin{equation}
\mitbf{R} = 
\begin{bmatrix}
\hat{\mitbf{v}} & \hat{\mitbf{w}} & -\hat{\mitbf{u}}
\end{bmatrix} \: .
\label{eq:R}
\end{equation}

\section{Anomalous quantities in gravity modeling}

@@

The difference between $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$, at the same 
point $P$, defines a quantity called \textit{gravity disturbance 
vector}, which is given by:
\begin{equation}
\delta \mitbf{g}_{P} = 
\mitbf{g}_{P} - \mitbf{\gamma}_{P} \: .
\label{eq:gravity-disturbance-vector}
\end{equation}
As the centrifugal part of the gravity vector and the normal gravity vector 
is equal to the centrifugal component of the gravity vector, 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
a purely gravitational (and consequently harmonic) quantity, which is 
caused by contrasts between the actual internal 
density distribution of the Earth and the (unknown) internal density 
distribution of the normal Earth.
In applied geophysics, these density differences are generally 
called \textit{anomalous masses} (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, 
\citeyear{forsberg1984}) or \textit{gravity sources} (e.g., 
\citeauthor{blakely1996}, \citeyear{blakely1996}). Here, we opted for 
using the last term.
Figure \ref{fig:surfaces} illustrates the gravity vector 
$\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$ and 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ at a point $P$ 
located on the Earth's surface.

The difference between gravity $g_{P}$ (magnitude of the gravity 
vector $\mitbf{g}_{P}$) and normal gravity $\gamma_{P}$ 
(magnitude of the normal gravity vector $\mitbf{\gamma}_{P}$), at the 
same point $P$, is called \textit{gravity disturbance} 
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005} 
and can be represented as follows:
\begin{equation}
\delta g_{P} = g_{P} - \gamma_{P} \: .
\label{eq:gravity-disturbance}
\end{equation}
Notice that the gravity disturbance $\delta g_{P}$ is not equivalent 
to the magnitude of the gravity disturbance vector 
$\delta \mitbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.
As properly pointed out by \citet{hackney-featherstone2003},
the gravity disturbance is a very-well established quantity in geodesy,
but appears to be less well known in geophysics.

%Consider a line (dashed line in Figure \ref{fig:surfaces}) passing 
%through a point $Q$ on the surface of the reference ellipsoid and 
%a point $P$ on the Earth's surface

The gravity anomaly is the commonly used quantity in applied 
geophysics. It is defined as the difference between the 
gravity at a point $Q^{\prime}$ on the geoid (a particular 
equipotential surface of the gravity potential) and the normal gravity 
at a point $Q$ on the ellipsoid surface,
both at the same geodetic latitude and longitude (Figure \ref{fig:surfaces}).
Notice that, by definition, the gravity anomaly depends on 
longitude and latitude only and is not a function of height 
\citep{barthelmes2013}. As a consequence, it is not possible to compute,
for example, the upward continuation of a gravity anomaly.
However, many authors in the literature have computed the upward 
continuation of gravity anomalies. All these authors have implicitly 
considered that the gravity anomaly is an approximation of 
the gravity disturbance, which in turn can be represented by a 
harmonic function that can be continued upward.

@@

\section{Mathematical description of the gravity disturbance in a local coordinate system}

In this coordinate system, the observed gravity vector
$\mitbf{g}_{i}$, at a point $(x_{i}, y_{i}, z_{i})$, 
$i = 1, ..., N$, can be represented by
\begin{equation}
\mitbf{g}_{i} = \mitbf{\gamma}_{i} + \Delta \mitbf{g}_{i} \: ,
\label{eq:gravity-vector}
\end{equation}
where $\mitbf{\gamma}_{i}$ and $\Delta \mitbf{g}_{i}$
are, respectively, the normal gravity vector and a
disturbing gravitational attraction produced by the anomalous 
masses at the point $(x_{i}, y_{i}, z_{i})$.

For each point $(x_{i}, y_{i}, z_{i})$ in the topocentric
Cartesian coordinate system (Figure \ref{fig:TCS}b), there is
a corresponding point $(h_{i}, \varphi_{i}, \lambda_{i})$
in the geodetic coordinate system (Figure \ref{fig:GCS-GGS}).

By definition, the gravity disturbance $\delta g_{i}$,
at the point $(x_{i}, y_{i}, z_{i})$, is given by 
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}:
\begin{equation}
\delta g_{i} =  g_{i} - \gamma_{i} \: ,
\label{eq:gravity-disturbance}
\end{equation}
where $g_{i} = \| \mitbf{g}_{i} \|$ and 
$\gamma_{i} = \| \mitbf{\gamma}_{i} \|$ are,
respectively, the observed gravity and the normal gravity at the
point $(x_{i}, y_{i}, z_{i})$.
Fortunately, the condition 
$\gamma_{i} \gg \| \Delta \mathbf{g}_{i} \|$ 
is met at all points located above or on the Earth's surface.
By combining this condition and the definition of observed gravity vector
(equation \ref{eq:gravity-vector}), we can approximate the observed gravity
$g_{i}$ by a first order Taylor's expansion as follows 
\citep{sanso_sideris2013}:
\begin{equation}
g_{i} \approx \gamma_{i} + 
\hat{\mitbf{\gamma}}_{i}^{\top} \Delta \mitbf{g}_{i} \: ,
\label{eq:gobs-approx}
\end{equation}
where $^{\top}$ denotes transposition,
$\hat{\mitbf{\gamma}}_{i} = -\hat{\mitbf{u}}_{i}$ is a unit 
vector with the same direction as the normal gravity vector 
$\mitbf{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$, in the
topocentric Cartesian coordinate system (Figure \ref{fig:TCS}b), and
$\hat{\mitbf{u}}_{i}$ is the unit vector $\hat{\mitbf{u}}$ 
(equation \ref{eq:unit-vectors}) evaluated at the corresponding 
point $(h_{i}, \varphi_{i}, \lambda_{i})$ in the 
geodetic coordinate system (Figure \ref{fig:GCS-GGS}).

This approximation, which is known in geodesy \citep[e.g.,][]{sanso_sideris2013},
is largely used in applied geophysics for representing
total-field anomalies \citep[e.g.,][]{blakely1996}.
Notice that, in local- or regional-gravity studies, the unit
vector $\hat{\mitbf{\gamma}}_{i}$ (equation \ref{eq:gobs-approx}) 
may be considered constant throughout the study area and 
parallel to the $z$ axis of the topocentric Cartesian coordinate 
system (Figure \ref{fig:TCS}b).
Consequently, by using the approximation defined in equation 
\ref{eq:gobs-approx}, the gravity disturbance 
(equation \ref{eq:gravity-disturbance}) can be rewritten as 
follows 
\begin{equation}
\delta g_{i} \approx \hat{\mitbf{\gamma}}_{P}^{\top} \Delta \mitbf{g}_{i} \: ,
\label{eq:gravity-disturbance-approx}
\end{equation}
where $\hat{\mitbf{\gamma}}_{P}$ represents the unit vector with
the same direction as the normal gravity vector $\mitbf{\gamma}_{P}$
at the origin $P$ of the topocentric Cartesian coordinate system
(Figure \ref{fig:TCS}).
This equation shows that the gravity disturbance $\delta g_{i}$ (equation \ref{eq:gravity-disturbance}) is different from the magnitude
of the disturbing gravitational attraction $\Delta \mitbf{g}_{i}$
produced by the gravity sources. Rather, it represents the 
component of the $\Delta \mitbf{g}_{i}$ on the direction of the normal
gravity vector \citep{sanso_sideris2013}. 
In the topocentric Cartesian coordinate system,
the gravity disturbance $\delta g_{i}$ (equation \ref{eq:gravity-disturbance-approx})
can be defined as the vertical component of the gravitational attraction exerted by
the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
As a consequence, the gravity disturbance produced by
a homogeneous gravity source with density contrast $\Delta\rho$
(in $kg / m^{3}$) can be represented by the following harmonic function:
\begin{equation}
d_{i} = k_{g} \, G \, \Delta\rho \, \partial_{z} \phi_{i} \: ,
\label{eq:gz-local}
\end{equation}
where $G$ is the Newtonian constant of gravitation
(in $m^{3} / (kg \, s^{2})$),
$k_{g} = 10^{5}$ is a constant factor
transforming from $m/s^{2}$ to milligal (mGal),
and $\partial_{z} \phi_{i}$ is a harmonic function representing the first 
derivative, evaluated at the observation point $(x_{i},y_{i},z_{i})$, 
$i = 1, ..., N$, of the function 
\begin{equation}
\phi(x,y,z) = \int\int\limits_{v}\int \frac{1}{\sqrt{(x - x^{\prime})^{2} + 
(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: dv
\label{eq:phi}
\end{equation}
with respect to the variable $z$. The 
integral is conducted over the coordinates $x^{\prime}$, $y^{\prime}$ 
and $z^{\prime}$ within the volume $v$ of the gravity source.
This equation can be easily generalized for the case
of multiple gravity sources.

Practically all the literature about gravity modeling
use the quantity $d_{i}$ (equation \ref{eq:gz-local})
to represent the gravity anomaly produced by 
gravity sources \citep[e.g.,][]{blakely1996}. 
Consequently, almost all the geophysicists use the gravity
anomaly as an approximation of the gravity disturbance
produced by the gravity sources. Notice that $d_{i}$ 
(equation \ref{eq:gz-local}) does not depend on the
geoidal surface. Rather, it depends on the relative position 
of the observation points $(x_{i}, y_{i}, z_{i})$ with respect to
the gravity sources in a topocentric Cartesian coordinate
system (Figure \ref{fig:TCS}b).

@@

We debate the conceptual differences between the gravity
disturbance and the gravity anomaly.
Our reasoning suggest that the gravity disturbance is the
more appropriated quantity for representing the gravity
effect produced by the gravity sources.
In summary, we point out that:

@@

\item Almost all interpretation techniques assume, implicitly or
directly, that the gravity data is harmonic (e.g., 
upward/downward continuation, 
data processing with equivalent layer,
conversions between gravity and magnetic data,
computation of vertical derivatives via Fourier and Hilbert
transforms). As a consequence, they implicitly or directly assume that
the gravity data approximates the gravitational disturbance.

\item Almost all forward modeling techniques compute
the vertical component of the gravitational attraction 
exerted by the geological bodies at the observation points.
Notice that the gravity anomaly requires the computation of gravity
on the Geoid, which is generally within the topographic masses.
Hence, almost all geophysicists implicitly compute 
the gravity disturbance.

>>>>

\title[Theoretical aspects in gravity modelling]
      {Theoretical aspects in gravity modelling}

\author[Oliveira Jr et al.]
{Vanderlei C. Oliveira Jr$^1$\thanks{Email: vanderlei@on.br}, 
Kristoffer A. T. Hallam$^{1}$, Val\'{e}ria C. F. Barbosa$^1$
and Leonardo Uieda$^2$ \\
$^1$ Department of Geophysics, Observat\'{o}rio Nacional, Rio de Janeiro, Brazil \\
$^2$ Department of Geology and Geophysics, University of Hawaii, Manoa, USA
}

@@

\begin{summary}
 The gravity anomaly has long been used by geophysicists for 
 representing the gravitational effect produced by gravity
 sources.This quantity, however, can be used in this way as a good
 approximation. Rigorously, the gravitational effect produced by the sources
 is represented by the gravity disturbance, which is a well-known 
 quantity in geodesy.
 In this paper, we discuss the fundamental concepts supporting 
 the fact that gravity disturbance is the more appropriated quantity
 to be used for estimating density variations related to geological
 bodies in subsurface. We stress that almost all authors in the 
 literature have implicitly used the gravity anomaly as an approximation 
 of the gravity disturbance, which can be represented by a harmonic 
 function and, consequently, satisfies the premise behind most of 
 the interpretation techniques used in geophysics (e.g., 
 upward/downward continuation, 
 data processing with equivalent layer,
 conversions between gravity and magnetic data,
 computation of vertical derivatives via Fourier and Hilbert
 transforms).
 Although the gravity anomaly can be considered a good approximation of the 
 gravity disturbance for most practical applications, 
 it is important to bear in mind not only the terminology 
 changes, but also the conceptual assumptions.

\end{summary}

\begin{keywords}
 potential fields -- gravity disturbance -- gravity anomaly -- gravity modelling.
\end{keywords}

@@

The resultant of gravitational force and centrifugal force acting 
on a body at rest on the Earth's surface is called 
\textit{gravity vector} and its intensity is called 
\textit{gravity} \citep{heiskanen-moritz1967,
hofmann-wellenhof-moritz2005}.
In the case of gravimetry on moving platforms (e.g., airplanes,
helicopters, marine vessels), there are additional
non-gravitational accelerations due to the vehicle motion, 
such as Coriolis acceleration and high-frequency vibrations 
\citep{glennie-etal2000,nabighian-etal2005-grav,baumann-etal2012}.
Geophysicists use gravity for estimating the Earth's 
internal density distribution whereas geodesists use
gravity to estimate the Geoid \citep{li2001}.
Hence, geophysicists are usually interested 
in the gravitational component of the observed gravity, 
which is produced by the Earth's internal 
density distribution.

The first step of the procedure for isolating this 
gravitational component consists in removing the non-gravitational 
effects due to the vehicle motion and also the time variations 
such as Earth tides, instrumental drift and barometric 
pressure changes, for example.
If these effects are properly removed, the resultant 
gravity data can be considered as the sum of a 
centrifugal component due to the Earth's rotation and
a gravitational component produced by the whole Earth's
internal density distribution.
The isolation of this particular gravitational component 
and its subsequent use for estimating density 
distributions related to geological structures in subsurface 
are the main goals in applied geophysics \citep{blakely1996}.
The computation of the gravitational effect produced by
the geological structures is commonly known in geophysics as 
\textit{gravity modelling}.
This term is different from \textit{gravity field modelling}, which
is commonly used in geodesy to denote the characterization of the 
gravity field in a local, regional or global scale.

@@

\section{Terrestrial Reference Systems used in gravity modelling}

For geophysical purposes, there are three important Terrestrial Reference Systems used in gravity modelling. 
They rotate with the Earth and are used for describing
positions and movements of objects on and close to the Earth’s surface
\citep{torge2012}.

The first one is a geocentric system of Cartesian coordinates having 
the $Z$-axis coincident with the mean Earth's rotational axis,
the $X$-axis pointing to the Greenwich meridian and the $Y$-axis
directed so as to obtain a right-handed system (Fig. \ref{fig:GCS-GGS}).
This reference system can be found in the literature with different names: Mean Terrestrial System \citep[e.g.,][]{soler1976}, Earth-fixed geocentric Cartesian system \citep[e.g.,][]{torge2012} 
or Earth-centered Earth-fixed system \citep[e.g.,][]{bouman_etal2013}, for example. Here, we opted for simply using the term Geocentric Cartesian System (GCS).

Other important reference system is a geocentric system of 
geodetic coordinates, which is defined by the reference ellipsoid 
used in the Normal Earth model \citep{heiskanen-moritz1967, soler1976, 
torge2012, bouman_etal2013}. 
In this coordinate system, the position of a point 
is defined by the \textit{geometric height} $h$, 
\textit{geodetic latitude} $\varphi$ and \textit{longitude} $\lambda$ (Fig. \ref{fig:GCS-GGS}).
For convenience, we call this system Geocentric Geodetic System
(GGS). 
At a given point $(h, \varphi, \lambda)$, there are three 
mutually-orthogonal unit vectors (Fig. \ref{fig:GCS-GGS}) given by \citep{soler1976}:
\begin{equation}
\hat{\mitbf{u}} = 
\begin{bmatrix}
\cos\varphi \, \cos\lambda \\
\cos\varphi \, \sin\lambda \\
\sin\varphi
\end{bmatrix}
\hat{\mitbf{v}} = 
\begin{bmatrix}
-\sin\varphi \, \cos\lambda \\
-\sin\varphi \, \sin\lambda \\
\cos\varphi
\end{bmatrix}
\hat{\mitbf{w}} = 
\begin{bmatrix}
-\sin\lambda \\
\cos\lambda \\
0
\end{bmatrix} .
\label{eq:unit-vectors}
\end{equation}
The required equations to convert coordinates $(h, \varphi, 
\lambda)$ referred to the GGS into coordinates $(X, Y, Z)$ referred to 
the GCS (Fig. \ref{fig:GCS-GGS}) and vice versa can be easily found in 
the literature \citep[e.g.,][]{heiskanen-moritz1967, torge2012, 
bouman_etal2013}.

In a local or regional study, geophysicists commonly use a topocentric 
Cartesian coordinate system (TCS) with origin at a point $P$ on or 
close to the Earth's surface and axes $x$, $y$ and $z$ (Fig. 
\ref{fig:TCS}). In the TCS, the axes $x$ and $y$ are parallel to 
the unit vectors $\hat{\mitbf{v}}_{P}$ $\hat{\mitbf{w}}_{P}$, respectively,
whereas the $z$-axis is opposite to the unit vector 
$\hat{\mitbf{u}}_{P}$ (eq. \ref{eq:unit-vectors}) and points downward.
Consider, for example, a TCS (Fig. \ref{fig:TCS}) with origin at a point $P$ with
coordinates $(X_{P}, Y_{P}, Z_{P})$ referred to the GCS (Fig. \ref{fig:GCS-GGS}).
In this case, the relationship between the coordinates $(X, Y, Z)$ of a 
point in the GCS and the coordinates $(x, y, z)$ of the same point in the 
TCS is given by:
\begin{equation}
\begin{bmatrix}
x \\
y \\
z 
\end{bmatrix} =
\mitbf{R}^{\top} \begin{bmatrix}
X - X_{P} \\
Y - Y_{P} \\
Z - Z_{P}
\end{bmatrix} \: ,
\label{eq:GCS2TCS}
\end{equation}
where $\mitbf{R}$ is an orthogonal matrix whose columns
are defined by the unit vectors $\hat{\mitbf{u}}$, $\hat{\mitbf{v}}$ and
$\hat{\mitbf{w}}$ (eq. \ref{eq:unit-vectors}), evaluated at the
point $P$, as follows:
\begin{equation}
\mitbf{R} = 
\begin{bmatrix}
\hat{\mitbf{v}} & \hat{\mitbf{w}} & -\hat{\mitbf{u}}
\end{bmatrix} \: .
\label{eq:R}
\end{equation}

\section{Anomalous quantities in gravity modelling}

@@

The difference between $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$, at the same 
point $P$, defines a quantity called \textit{gravity disturbance 
vector}, which is given by:
\begin{equation}
\delta \mitbf{g}_{P} = 
\mitbf{g}_{P} - \mitbf{\gamma}_{P} \: .
\label{eq:gravity-disturbance-vector}
\end{equation}
As the centrifugal part of the gravity vector and the normal gravity vector 
is equal to the centrifugal component of the gravity vector, 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
a purely gravitational (and consequently harmonic) quantity, which is 
caused by contrasts between the actual internal 
density distribution of the Earth and the (unknown) internal density 
distribution of the normal Earth.
In applied geophysics, these density differences are generally 
called \textit{anomalous masses} (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, 
\citeyear{forsberg1984}) or \textit{gravity sources} (e.g., 
\citeauthor{blakely1996}, \citeyear{blakely1996}). Here, we opted for 
using the last term.
Fig. \ref{fig:surfaces} illustrates the gravity vector 
$\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$ and 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ at a point $P$ 
located on the Earth's surface.

The difference between gravity $g_{P}$ (magnitude of the gravity 
vector $\mitbf{g}_{P}$) and normal gravity $\gamma_{P}$ 
(magnitude of the normal gravity vector $\mitbf{\gamma}_{P}$), at the 
same point $P$, is called \textit{gravity disturbance} 
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005} 
and can be represented as follows:
\begin{equation}
\delta g_{P} = g_{P} - \gamma_{P} \: ,
\label{eq:gravity-disturbance}
\end{equation}
where $g_{P} = \| \mitbf{g}_{P} \|$ and 
$\gamma_{P} = \| \mitbf{\gamma}_{P} \|$ are,
respectively, the gravity and the normal gravity at the
point $P$.
Notice that the gravity disturbance $\delta g_{P}$ is not equivalent 
to the magnitude of the gravity disturbance vector 
$\delta \mitbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.
As properly pointed out by \citet{hackney-featherstone2003},
the gravity disturbance is a very-well established quantity in geodesy,
but appears to be less well known in geophysics.

The gravity anomaly is the commonly used quantity in applied 
geophysics. It is defined as the difference between the 
gravity at a point $Q^{\prime}$ on the Geoid (a particular 
equipotential surface of the gravity potential) and the normal gravity 
at a point $Q$ on the ellipsoid surface,
both at the same geodetic latitude and longitude (Fig. \ref{fig:surfaces}).
Notice that, by definition, the gravity anomaly depends on 
longitude and latitude only and is not a function of height 
\citep{barthelmes2013}. As a consequence, it is not possible to compute,
for example, the upward continuation of a gravity anomaly.
However, many authors in the literature have computed the upward 
continuation of gravity anomalies. All these authors have implicitly 
considered that the gravity anomaly is an approximation of 
the gravity disturbance, which in turn can be represented by a 
harmonic function that can be continued upward.

@@

\section{Mathematical description of the gravity disturbance at the TCS}

From eq. \ref{eq:gravity-disturbance-vector}, the gravity vector $\mitbf{g}_{i}$, at a point 
$(x_{i}, y_{i}, z_{i})$ in the TCS (Fig. \ref{fig:TCS}b), can be represented by:
\begin{equation}
\mitbf{g}_{i} = \mitbf{\gamma}_{i} + \delta \mitbf{g}_{i} \: ,
\label{eq:gravity-vector-TCS}
\end{equation}
where $\mitbf{\gamma}_{i}$ and $\delta \mitbf{g}_{i}$
are, respectively, the normal gravity vector and the 
gravity disturbance vector produced by the anomalous 
masses at the point $(x_{i}, y_{i}, z_{i})$.

The gravity $g_{i}$ can be approximated by a first order Taylor's 
expansion as follows \citep{sanso_sideris2013}:
\begin{equation}
g_{i} \approx \gamma_{i} + 
\hat{\mitbf{\gamma}}_{i}^{\top} \delta \mitbf{g}_{i} \: ,
\label{eq:gobs-approx}
\end{equation}
where $^{\top}$ denotes transposition,
$\hat{\mitbf{\gamma}}_{i} = -\hat{\mitbf{u}}_{i}$ is a unit 
vector with the same direction as the normal gravity vector 
$\mitbf{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$, in the
topocentric Cartesian coordinate system (Fig. \ref{fig:TCS}b), and
$\hat{\mitbf{u}}_{i}$ is the unit vector $\hat{\mitbf{u}}$ 
(eq. \ref{eq:unit-vectors}) evaluated at the corresponding 
point $(h_{i}, \varphi_{i}, \lambda_{i})$ in the 
GGS (Fig. \ref{fig:GCS-GGS}).
This approximation can be made because, fortunately, 
the condition $\gamma_{i} \gg \| \delta \mathbf{g}_{i} \|$ 
is met at all points located above or on the Earth's surface.
This approximation, which is known in geodesy \citep[e.g.,][]{sanso_sideris2013},
is largely used in applied geophysics for representing
total-field anomalies \citep[e.g.,][]{blakely1996}.
Notice that, in local- or regional-gravity studies, the unit
vector $\hat{\mitbf{\gamma}}_{i}$ (eq. \ref{eq:gobs-approx}) 
may be considered constant throughout the study area and 
parallel to the $z$ axis of the TCS (Fig. \ref{fig:TCS}b).
By using the approximation defined in eq. 
\ref{eq:gobs-approx}, the gravity disturbance 
(eq. \ref{eq:gravity-disturbance}) can be rewritten as 
follows 
\begin{equation}
\delta g_{i} \approx \hat{\mitbf{\gamma}}_{P}^{\top} \delta \mitbf{g}_{i} \: ,
\label{eq:gravity-disturbance-approx}
\end{equation}
where $\hat{\mitbf{\gamma}}_{P}$ represents the unit vector with
the same direction as the normal gravity vector $\mitbf{\gamma}_{P}$
at the origin $P$ of the TCS (Fig. \ref{fig:TCS}b).
This equation shows that the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance}) is different from the magnitude
of the gravity disturbance vector $\delta \mitbf{g}_{i}$
produced by the gravity sources. Rather, it represents the 
component of $\delta \mitbf{g}_{i}$ on the direction of the normal
gravity vector \citep{sanso_sideris2013}.

In the TCS (Fig. \ref{fig:TCS}b),
the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance-approx})
can be defined as the $z$-component of the gravitational attraction exerted by
the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
As a consequence, the gravity disturbance produced by
a gravity source can be represented by the following harmonic function:
\begin{equation}
d_{i} = \partial_{z} \phi_{i} \: ,
\label{eq:gz-local}
\end{equation}
where $\partial_{z} \phi_{i}$ is the first 
derivative, evaluated at the point $(x_{i},y_{i},z_{i})$, of the function 
\begin{equation}
\phi(x,y,z) = \int\int\limits_{v}\int 
\frac{c_{g} \, G \, \Delta\rho(x^{\prime}, y^{\prime}, z^{\prime}) \: dv}
{\sqrt{(x - x^{\prime})^{2} + 
(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: ,
\label{eq:phi}
\end{equation}
with respect to the variable $z$. Is this integral, 
$G$ is the Newtonian constant of gravitation (in $m^{3} kg^{-1} s^{-2}$),
$c_{g} = 10^{5}$ is a constant factor transforming from $m s^{-2}$ 
to milligal (mGal), $\Delta\rho(x^{\prime}, y^{\prime}, z^{\prime})$
is the density contrast (in $kg / m^{3}$) at a point 
$(x^{\prime}, y^{\prime}, z^{\prime})$ within 
the volume $v$ of the gravity source and the integration is conducted 
over $x^{\prime}$, $y^{\prime}$ and $z^{\prime}$.
This equation can be easily generalized for the case
of multiple gravity sources.

Practically all the literature about gravity modelling
use the quantity $d_{i}$ (eq. \ref{eq:gz-local})
to represent the gravity anomaly produced by 
gravity sources \citep[e.g.,][]{blakely1996}. 
Consequently, almost all geophysicists use the gravity
anomaly as an approximation of the gravity disturbance
produced by the gravity sources. Notice that $d_{i}$ 
(eq. \ref{eq:gz-local}) does not depend on the
height with respect to the Geoid (orthometric height). 
Rather, it depends on the relative position 
of the observation points $(x_{i}, y_{i}, z_{i})$ with respect to
the gravity sources in the TCS (Fig. \ref{fig:TCS}b).
It is important to notice that, contrary to what is written in 
some basic texts, gravimeters do not measure the quantity
$d_{i}$ (eq. \ref{eq:gz-local}).

@@

We debate the conceptual differences between the gravity
disturbance and the gravity anomaly.
Our reasoning suggests that the gravity disturbance is the
more appropriated quantity for representing the gravity
effect produced by the gravity sources.
In summary, we point out that:

@@

\item Almost all interpretation techniques assume, implicitly or
directly, that the gravity data is harmonic (e.g., 
upward/downward continuation, 
data processing with equivalent layer,
conversions between gravity and magnetic data,
computation of vertical derivatives via Fourier and Hilbert
transforms). As a consequence, they implicitly or directly assume that
the gravity data approximate the gravitational disturbance.

\item Almost all forward modelling techniques compute
the vertical component of the gravitational attraction 
exerted by the gravity sources at the observation points.
Notice that the gravity anomaly requires the computation of gravity
on the Geoid, which is generally within the topographic masses.
Hence, almost all forward modelling techniques compute 
the gravity disturbance.
